---
title: 'LESSON TITLE'
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
editor_options: 
  chunk_output_type: inline
---


```{r, echo = F, message = F, warning = F}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr, here, reactable, gt, flextable)

## functions
source(here::here("global/functions/lesson_functions.R"))
```

# Introduction {.unnumbered}

[Few sentences describing the main topic of the lesson and why it's important. Nice to include an image here too.]

# Learning Objectives {.unnumbered}

-   XXXX

-   XXXX

-   XXXX

-   XXXX

# Packages

This lesson will require the following packages to be installed and loaded:

```{r warning = F, message = F, echo = F}
# Load packages 
if(!require(pacman)) install.packages("pacman")
#pacman::p_load(INSERT PACKAGES REQUIRED HERE)

```

------------------------------------------------------------------------

# Introduction to the Dataset 

-   XXXX

------------------------------------------------------------------------

# [RELEVANT LESSON SECTION NAME]

-   XXXX

------------------------------------------------------------------------

# [RELEVANT LESSON SECTION NAME]

-   XXXX

------------------------------------------------------------------------

# [RELEVANT LESSON SECTION NAME]

-   XXXX

------------------------------------------------------------------------

::: demo

[XXX DEMO HERE XXX]

::: 

------------------------------------------------------------------------

::: reminder

[XXX SIDE NOTES HERE XXXX]

:::

------------------------------------------------------------------------

::: recap

[XXX RECAP HERE XXX]

:::

------------------------------------------------------------------------

::: practice

[XXX PRACTICE QUESTIONS HERE XXX] 
- NOTE: Make sure to add the answers to all practice at the end of the lesson! We are no longer using autograders for the GF project.

:::

------------------------------------------------------------------------

::: r-practice

[XXX PRACTICE QUESTIONS HERE XXX] 
- NOTE: Make sure to add the answers to all practice at the end of the lesson! We are no longer using autograders for the GF project.

:::

------------------------------------------------------------------------

::: challenge

[XXX CHALLENGE PRACTICE QUESTIONS HERE XXX]
- NOTE: Make sure to add the answers to all practice at the end of the lesson! We are no longer using autograders for the GF project.

::: 

------------------------------------------------------------------------

::: pro-tip

[XXX PRO-TIPS HERE XXX]

:::

------------------------------------------------------------------------

::: vocab

[XXX RELEVANT VOCUBULARY WORDS HERE XXX]

:::

------------------------------------------------------------------------

::: side-note

[XXX SIDE NOTES HERE XXXX]

:::

------------------------------------------------------------------------

::: watch-out

[XXX TIPS HERE XXX]

:::

------------------------------------------------------------------------

::: key-point

[XXX KEY POINTS HERE XXX]

:::

------------------------------------------------------------------------

::: error

[XXX ERRORS HERE XXX]

::: 

------------------------------------------------------------------------

::: rstudio-cloud

[XXX RSTUDIO CLOUD TIPS HERE XXX]

::: 

------------------------------------------------------------------------

# WRAP UP! {.unnumbered}

[XXX NICE WRAP UP MESSAGE OR SUMMARY HERE XXX]

------------------------------------------------------------------------

# Answer Key {.unnumbered}

------------------------------------------------------------------------

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r .tgc_contributors_list(ids = c("kendavidn", "joy"))` 
(make sure to update the contributor list accordingly!)




## References {.unlisted .unnumbered}

-   XXXX


<!-- (Chicago format. You can use https://www.citationmachine.net) -->


`r .tgc_license()`
