---
title: 'T-tests for Comparing Means: Independent Samples, Paired Samples, and Effect Sizes'
format:
  html:
    embed-resources: true
---

```{r, echo = F, message = F, warning = F}
# Load packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr, here, reactable, gt, flextable, effectsize)

# Source functions
source(here("global/functions/lesson_functions.R"))

# Knitr options
knitr::opts_chunk$set(
  warning = F, message = F,
  class.source = "tgc-code-block", error = T
)

# Set ggplot theme
theme_set(theme_minimal())
```

------------------------------------------------------------------------

# Introduction {.unnumbered}

In this lesson, we learn how to use **t-tests** to compare the means of two groups. We’ll walk through **independent samples t-tests**, **paired samples t-tests**, and how to interpret **effect sizes** like **Cohen’s d**.

# Learning Objectives {.unnumbered}

By the end of this lesson, you will be able to:

-   Understand what a t-test is and when to use it
-   Distinguish between independent and paired t-tests
-   Conduct t-tests in R using real data
-   Interpret the results, including p-values and effect sizes

------------------------------------------------------------------------

# What is a t-test?

A **t-test** is a statistical test used to compare the **means** of **two groups**. We use it when we want to know:

> **Is the difference between two sample means likely to reflect a real difference in the population?**

The t-test works by comparing the observed difference in sample means to what we'd expect by random chance.

There are two main types of t-tests:

| Test Type | Use When... |
|---------------------------------|---------------------------------------|
| **Independent Samples t-test** | You're comparing **two different groups** (e.g. men vs. women) |
| **Paired Samples t-test** | You're comparing **the same group at two times** (e.g. before and after treatment) |

### Where does the "t" come from?

The "t" in t-test refers to the **t-distribution**, which is a probability distribution similar to the normal distribution, but with fatter tails. We use the t-distribution instead of the normal distribution when working with **small samples** or when **population standard deviation is unknown**.

The basic t-statistic formula is:

$$
t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
$$

Where:

-   ( \bar{x}\_1, \bar{x}\_2 ): sample means of the two groups
-   ( s_1\^2, s_2\^2 ): sample variances
-   ( n_1, n_2 ): sample sizes

------------------------------------------------------------------------

# Packages

Let’s load the packages needed for this lesson:

```{r}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(
  tidyverse,
  here,
  janitor,
  effectsize
)
```

------------------------------------------------------------------------

# The Diabetes in China Dataset

As in the previous lesson, we will use the dataset from the Chinese diabetes cohort study. Let’s load and inspect the dataset:

```{r}
diabetes_china <- read_csv(here("data/diabetes_china_chen.csv"))
```

```{r render = .reactable_5_rows, echo = F}
head(diabetes_china)
```

## Recap: BMI Categories

We will categorize participants into two groups based on BMI:

```{r}
diabetes_china <- diabetes_china %>% 
  mutate(bmi_cat = case_when(
    bmi < 25 ~ 1,
    bmi >= 25 ~ 2
  ))
```

-   `1` = BMI \< 25 (Normal weight)
-   `2` = BMI ≥ 25 (Overweight/obese)

------------------------------------------------------------------------

# Independent Samples t-test

Let’s test whether people with BMI \< 25 have different **systolic blood pressure (SBP)** compared to those with BMI ≥ 25.

```{r}
t_test_independent <- t.test(sbp_mm_hg ~ bmi_cat, data = diabetes_china)
t_test_independent
```

### Interpretation

-   The p-value tells us whether the difference in mean SBP between the two BMI groups is **statistically significant**.

-   The means of both groups and their confidence intervals will help us understand the direction and magnitude of the difference.

-   If **p \< 0.05**, we conclude the difference is **statistically significant**.

-   The **mean systolic blood pressure** (SBP) in the **BMI \< 25 group** was **116.07 mm Hg**, while in the **BMI ≥ 25 group** it was **126.49 mm Hg**.

-   The **difference of \~10.4 mm Hg** is statistically significant, with a **p-value \< 2.2e-16**.

-   The 95% confidence interval for the difference in means is **\[-10.57, -10.27\]**, which does **not** include 0 — another sign of a significant result.

> ✅ The difference in SBP between BMI groups is **statistically significant**.

------------------------------------------------------------------------

# Paired Samples t-test

Sometimes, we measure the **same people** twice. For example, let’s imagine we want to test whether **fpg_mmol_l** (fasting plasma glucose) changed over time, from baseline to final follow-up. We'll use the variables `fpg_mmol_l` (baseline) and `fpg_of_final_visit_mmol_l` (follow-up).

```{r}
# Remove rows with missing values in either time point
paired_data <- diabetes_china %>% 
  filter(!is.na(fpg_mmol_l), !is.na(fpg_of_final_visit_mmol_l))

t_test_paired <- t.test(paired_data$fpg_mmol_l, paired_data$fpg_of_final_visit_mmol_l, paired = TRUE)
t_test_paired
```

### Interpretation

-   A paired t-test accounts for the **within-subject** variability.

-   If **p \< 0.05**, we conclude that there was a significant change in fasting glucose over time.

-   The **mean fasting plasma glucose (FPG)** decreased by **0.218 mmol/L** from baseline to follow-up.

-   The paired t-test shows a **p-value \< 2.2e-16**, indicating the difference is **statistically significant**.

-   The 95% confidence interval is **\[-0.221, -0.215\]**, meaning we can be confident the true mean decrease lies within this range.

> ✅ FPG significantly decreased over time within individuals.

------------------------------------------------------------------------

# Effect Sizes: How Big is the Difference?

Even if the t-test gives a statistically significant result, we want to know if the difference is **meaningful**. That’s what **effect sizes** tell us.

### Cohen’s d

Cohen's d quantifies the difference between two means in **standard deviation units**.

For **independent samples**:

```{r}
cohen_d_independent <- cohens_d(sbp_mm_hg ~ as.factor(bmi_cat), data = diabetes_china)
cohen_d_independent
```

For **paired samples**:

```{r}
cohen_d_paired <- cohens_d(paired_data$fpg_mmol_l, paired_data$fpg_of_final_visit_mmol_l, paired = TRUE)
cohen_d_paired
```

### Interpretation of Cohen's d

| Cohen’s d value | Interpretation |
|-----------------|----------------|
| \~0.2           | Small effect   |
| \~0.5           | Medium effect  |
| \~0.8+          | Large effect   |

#### Independent Samples

-   **Cohen’s d = -0.66**, which indicates a **medium-to-large effect size**.
-   This tells us the **difference in SBP between BMI groups** is not just statistically significant — it is also **practically meaningful**.

#### Paired Samples

-   **Cohen’s d = -0.32**, which is considered a **small-to-medium effect size**.
-   The decrease in FPG is statistically reliable, but the **magnitude of change is modest** — it may or may not be meaningful in a clinical sense depending on the context.

### Summary Table

| Comparison | p-value | Cohen’s d | Interpretation |
|--------------------|---------------|---------------|-----------------------|
| SBP by BMI group (Independent) | \< 2.2e-16 | 0.66 | Statistically significant and **meaningful** |
| FPG baseline vs. follow-up (Paired) | \< 2.2e-16 | 0.32 | Statistically significant, **small effect** |

✅ **Statistically significant** differences in both cases\
✅ **Effect size** suggests the SBP difference is practically important, while the FPG change is smaller

# Visualizing the Results

### SBP by BMI Category

```{r}
ggplot(diabetes_china, aes(x = as.factor(bmi_cat), y = sbp_mm_hg)) +
  geom_boxplot(fill = "skyblue") +
  labs(x = "BMI Category (1 = <25, 2 = ≥25)", y = "Systolic BP (mm Hg)",
       title = "Systolic Blood Pressure by BMI Category") +
  theme_minimal()
```

### Fasting Glucose Before and After (Paired)

```{r eval =FALSE}
paired_data_long <- paired_data %>%
  pivot_longer(cols = c(fpg_mmol_l, fpg_of_final_visit_mmol_l),
               names_to = "time", values_to = "fpg") %>%
  mutate(time = recode(time,
                       "fpg_mmol_l" = "Baseline",
                       "fpg_of_final_visit_mmol_l" = "Follow-up"))

ggplot(paired_data_long, aes(x = time, y = fpg, group = id)) +
  geom_line(alpha = 0.1) +
  geom_boxplot(aes(group = time), width = 0.2, color = "darkblue") +
  labs(x = "", y = "FPG (mmol/L)",
       title = "Fasting Plasma Glucose: Baseline vs. Follow-up") +
  theme_minimal()
```

------------------------------------------------------------------------

# Summary Table

| Comparison | p-value | Cohen’s d | Interpretation |
|------------------|-----------------|-----------------|----------------------|
| SBP by BMI group (Independent) | *(insert here)* | *(insert here)* | *(statistical + practical meaning)* |
| FPG baseline vs. follow-up (Paired) | *(insert here)* | *(insert here)* | *(statistical + practical meaning)* |

✅ Use p-values to check for **statistical significance**\
✅ Use effect size to assess **practical significance**

------------------------------------------------------------------------

# Assumptions of the t-test

For your t-test results to be **valid**, certain assumptions must be met. These depend on whether you are using an **independent samples** t-test or a **paired samples** t-test.

## 1. Independent Samples t-test Assumptions

| Assumption | What It Means | How to Check It in R |
|-----------------|----------------------------------|---------------------|
| **Independence of observations** | Each observation should be from a different person or unit | Based on study design (not testable in code) |
| **Normality** | The outcome variable should be approximately normally distributed **in each group** | Use `ggplot` or `shapiro.test()` per group |
| **Equal variances** (optional) | The variance in each group is roughly equal (only required for **Student's t-test**) | Use `var.test()` or `leveneTest()` |

Note: The **Welch t-test** (which is the default in `t.test()`) does **not require equal variances**, so it's generally safer.

## Checking for normality

Before running a t-test, one assumption we often check is **normality**—whether the data (or differences, in the case of paired tests) are roughly normally distributed. This matters most when sample sizes are **small** (e.g., \< 30 per group).

### 🧪 Checking normality with histograms or density plots

```{r}
# Density plots of SBP by BMI category
ggplot(diabetes_china, aes(x = sbp_mm_hg, fill = as.factor(bmi_cat))) +
  geom_density(alpha = 0.5) +
  labs(x = "Systolic Blood Pressure", fill = "BMI Category") +
  theme_minimal()
```

### 🧪 Shapiro-Wilk test for normality

The **Shapiro-Wilk test** is a formal test for normality. It works by comparing the shape of your data's distribution to a perfect normal distribution.

-   **Null hypothesis (H₀)**: The data **are normally distributed**\
-   **Alternative hypothesis (H₁)**: The data **are not normally distributed**

If the p-value is **less than 0.05**, we reject the null and conclude the data **deviate from normality**.

> ⚠️ However, for **large samples**, even tiny deviations from normality will result in a significant p-value—so this test may say the data aren't normal even when the departure is trivial. In those cases, it's more helpful to **look at plots** (histograms, Q-Q plots, or density curves).

The `shapiro.test()` function in R has a sample size limit of 5000, which makes it unsuitable for testing normality in very large datasets like yours (210,000+ rows).

We can test a random subset of your data (e.g., 3000–5000 rows per group), which is often enough to assess normality.

```{r}
# Take random samples of each BMI group (max 5000 rows)
set.seed(123)  # for reproducibility
group1_sample <- diabetes_china %>%
  filter(bmi_cat == 1, !is.na(sbp_mm_hg)) %>%
  slice_sample(n = 5000)

group2_sample <- diabetes_china %>%
  filter(bmi_cat == 2, !is.na(sbp_mm_hg)) %>%
  slice_sample(n = 5000)

# Shapiro-Wilk tests on the samples
shapiro.test(group1_sample$sbp_mm_hg)
shapiro.test(group2_sample$sbp_mm_hg)
```

Both groups have **p-values \< 0.05**, which means the Shapiro-Wilk test finds **evidence against normality** in both BMI groups' systolic blood pressure data.

However, since your sample size is **very large (N \> 100,000)**, this result is **expected**. Even slight deviations from a perfect bell curve will produce a significant result. That doesn't necessarily mean the data are unsuitable for a t-test.

> ✅ **Conclusion**: The formal test suggests non-normality, but with such a large sample, the **t-test is robust** to these violations. You can safely proceed with the t-test, especially since you're using the **Welch version**, which is also robust to unequal variances.

Still, it's a good idea to include a **density plot or Q-Q plot** to visually assess whether the distribution looks "normal enough."

> ⚠️ For large samples (like this one), the Shapiro test will almost always give **p \< 0.05**, suggesting non-normality — even when it doesn't matter much. Visual inspection (plots) is often more useful here.

------------------------------------------------------------------------

## 2. Paired Samples t-test Assumptions

| Assumption | What It Means | How to Check It in R |
|-----------------|------------------------------------|--------------------|
| **Dependent observations** | You are comparing two **related** measurements (e.g. same individual at two times) | Confirm with study design |
| **Normality of differences** | The **differences between paired values** should be normally distributed | Check the distribution of differences |

### 🧪 Check normality of differences

Take a random sample (e.g. 5000 rows) from the paired differences:

```{r}
set.seed(123)  # For reproducibility

# Sample 5000 rows for Shapiro-Wilk test
fpg_sample <- paired_data %>%
  drop_na(fpg_diff) %>%
  slice_sample(n = 5000)

shapiro.test(fpg_sample$fpg_diff)
```

#### Interpretation:

-   The **W statistic** (0.883) is quite far from 1, which indicates **deviation from normality**.

-   The **p-value is \< 2.2e-16**, which is much smaller than the common threshold of 0.05.

-   Therefore, we **reject the null hypothesis** that the differences are normally distributed.

>  ⚠️ **Conclusion:** The distribution of differences in fasting glucose **is not normal** based on this test.

#### But should we worry?

-   **Probably not**. You have **over 200,000 observations**, and in large samples, **even very minor skew or kurtosis** will lead to a "significant" Shapiro test.

-   More important is whether the distribution is **severely non-normal**, which you can assess visually using the density plot or histogram.

-   The t-test is **robust to violations of normality** when sample size is large (Central Limit Theorem).

### ✅ Final Recommendation:

> Although the Shapiro-Wilk test indicates non-normality, given the **very large sample size**, the **paired t-test is still valid**.\
> If the visual inspection shows no extreme skew or outliers, you can safely proceed with your t-test results.

------------------------------------------------------------------------

## What if assumptions are violated?

-   If the **normality assumption** is violated but your sample size is **large**, the t-test is generally **robust** (still valid).
-   If sample size is **small and data are skewed**, consider using a **non-parametric test**:
    -   **Independent samples** → Use `wilcox.test(x ~ group)`
    -   **Paired samples** → Use `wilcox.test(x, y, paired = TRUE)`

```{r}
# Example: non-parametric alternative to paired t-test
wilcox.test(paired_data$fpg_mmol_l, paired_data$fpg_of_final_visit_mmol_l, paired = TRUE)
```

> ✅ Best practice: always **check your assumptions**, but don't panic about small violations — especially with large samples.

------------------------------------------------------------------------

# Final Tips

-   Use **independent t-tests** when comparing **two different groups**
-   Use **paired t-tests** when comparing **the same group over time**
-   Always report both the **p-value** and the **effect size**
-   Visualize your data to help interpret results

------------------------------------------------------------------------

# Key Takeaways {.unnumbered}

Continue exploring these concepts with real-world data to gain deeper insight into public health analyses.

------------------------------------------------------------------------

# Answer Key {.unlisted .unnumbered}

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r .tgc_contributors_list(ids = c("joy", "kendavidn"))`

# References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Irizarry, Rafael A. 2019. *Introduction to Data Science: Statistics and Prediction Algorithms Through Case Studies*. <https://rafalab.dfci.harvard.edu/dsbook-part-2/>.
-   Cannell, Brad, and Melvin Livingston. n.d. *R for Epidemiology*. <https://www.r4epi.com/>.
-   GeeksforGeeks. 2025. "Introduction of Statistics and Its Types." <https://www.geeksforgeeks.org/introduction-of-statistics-and-its-types/>.
-   Starmer, Josh. *StatQuest with Josh Starmer* \[YouTube Channel\]. <https://www.youtube.com/channel/UCtYLUTtgS3k1Fg4y5tAhLbw>.
-   The GRAPH Courses. *FoSSA: Fundamentals of Statistical Software & Analysis*. <https://thegraphcourses.org/courses/fossa/>.
-   University of Miami Libraries. *Introduction to Data Analysis and R*. <https://www.library.miami.edu/data-services/data-analysis.html>.

`r .tgc_license()`
